---
title: "Petrographer Model Zoo with pins"
author: "petrographer team"
format:
  html:
    code-fold: true
toc: true
editor: visual
---

```{r}
#| label: setup
#| message: false
#| warning: false
library(petrographer)
# pins is optional; load if available for richer printing in this notebook
if (requireNamespace("pins", quietly = TRUE)) {
  library(pins)
}
```

## Overview

This notebook shows how to create and maintain a simple "model zoo" using the pins package:

- Configure a board (local folder, shared folder, S3, or Posit Connect)
- Pin trained petrographer models and their metadata
- Discover available models and versions
- Load models from pins for prediction
- Housekeeping: delete old pins

pins saves versioned artifacts and handles caching. For large files (> 500MB) prefer boards like S3 or a shared drive.

## Configure a Board

Choose a board based on how you share models:

```{r}
# Local board (per-machine, good for experiments)
board_local <- pg_board()
board_local
```

```{r}
# Shared folder (Dropbox / network drive)
# Uncomment and set a path visible to your team
# board_shared <- pins::board_folder("~/Dropbox/petrographer-models", versioned = TRUE)
```

```{r}
# Amazon S3 (requires credentials via env vars)
# board_s3 <- pins::board_s3(bucket = Sys.getenv("PETRO_S3_BUCKET"), prefix = "petrographer", versioned = TRUE)
```

```{r}
# Posit Connect (enterprise sharing)
# board_connect <- pins::board_connect()
```

## Pin a Trained Model

Pin a model directory that contains `model_final.pth` and `config.yaml` (created by `train_model()`).

```{r}
# Example: pin an existing trained model directory
example_model_dir <- "Detectron2_Models/shell_detector_v3"  # adjust to your model
if (dir.exists(example_model_dir)) {
  meta <- list(
    notes = "Pinned from notebook",
    owner = Sys.info()[["user"]]
  )
  publish_model(
    model_dir = example_model_dir,
    name = "shell_detector_v3",
    board = board_local,
    metadata = meta,
    include_metrics = TRUE
  )
}
```

Pin automatically right after training by passing `publish_after_train = TRUE` to `train_model()`:

```{r}
# if (FALSE) {
#   train_model(
#     data_dir = "data/processed/shell_mixed",
#     output_name = "shell_detector_v4",
#     num_classes = 5,
#     gpus = 1,
#     publish_after_train = TRUE,
#     model_board = board_local
#   )
# }
```

## Discover Models

List pins on the board and inspect metadata:

```{r}
available <- if (requireNamespace("pins", quietly = TRUE)) pins::pin_list(board_local) else character()
available
```

```{r}
if (requireNamespace("pins", quietly = TRUE) && length(available)) {
  info <- pins::pin_meta(board_local, available[[1]])
  str(info$user)
}
```

Search by text:

```{r}
if (requireNamespace("pins", quietly = TRUE)) {
  res <- pins::pin_search(board_local, "shell")
  if (!is.null(res) && "name" %in% names(res)) unique(res$name) else character()
}
```

## Versioning

Most boards support versions. With `pins`, each write creates a new version (when versioning is enabled on the board).

```{r}
if (requireNamespace("pins", quietly = TRUE) && length(available)) {
  pins::pin_versions(board_local, available[[1]])
}
```

## Load Models from Pins

Load a pinned model either explicitly or via `load_model()` auto-detection.

```{r}
if (length(available)) {
  gm <- get_model(available[[1]], board = board_local)
  mdl <- load_model(model_name = available[[1]], confidence = 0.5, device = "cpu")
}
```

## Use the Model

```{r}
# Example prediction (adjust the path to an existing image)
# if (exists("mdl") && file.exists("data/raw/sample.jpg")) {
#   res <- predict_image("data/raw/sample.jpg", mdl, use_slicing = TRUE)
#   head(res)
# }
```

## Housekeeping

Delete a pin (careful: deletes all versions on boards that don’t isolate versions):

```{r}
if (FALSE && requireNamespace("pins", quietly = TRUE) && length(available) > 0) {
  pins::pin_delete(board_local, available[[1]])
}
```

## Tips

- Prefer versioned boards (`versioned = TRUE`) so you can reproduce past runs.
- For large models or many versions, consider S3 instead of folder boards.
- Store concise training metadata with `publish_model()` to make discovery easier.
