---
title: "Image Preparation"
author: "Griffin Pitts and Nick Gauthier"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(here)
library(OpenImageR)
library(SuperpixelImageSegmentation)
library(jpeg)
```


This notebook reads in the ceramic thin section images from Lyons 2020. 

## Download

Check if the image data are in the local directory. If not, download them from Dropbox.
```{r}
data_dir <- here('data' , 'Lyons_data')

if(!file.exists(data_dir)){
  options(timeout = 180) # make this number bigger if your internet is slow and you get an error
  download.file('https://www.dropbox.com/sh/e4r2fnfq3gspytk/AACGGyuKY5xgKv3UAVWdXSG-a?dl=1',
                data_dir)
}
```

## Import

Read in a sample image.
```{r}
path <- here(data_dir, 'a', '15-604_xp_000013.jpg')
im <- readImage(path)
```

Check its dimensions.
```{r}
dim(im)
```

This opens up an interactive viewer.
```{r}
#imageShow(path)
```

Alternative using the `jpeg` package.

```{r}
jj <- readJPEG(path, native = TRUE)

plot(0:1,0:1,type="n",ann=FALSE,axes=FALSE) ; rasterImage(jj,0,0,1,1)
```

## Segmentation

This example is taken from [Image segmentation based on Superpixels and Clustering](https://mlampros.github.io/OpenImageR/articles/Image_segmentation_superpixels_clustering.html) vignette on the `OpenImageR` website.

```{r}
res_slic <- superpixels(input_image = im,
                       method = "slic",
                       superpixel = 200, 
                       compactness = 20,
                       return_slic_data = TRUE,
                       return_labels = TRUE, 
                       write_slic = "", 
                       verbose = TRUE)

res_slico <- superpixels(input_image = im,
                        method = "slico",
                        superpixel = 200, 
                        return_slic_data = TRUE,
                        return_labels = TRUE, 
                        write_slic = "", 
                        verbose = TRUE)
```
```{r}
par(mfrow=c(1,2), mar = c(0.2, 0.2, 0.2, 0.2))

plot_slic = OpenImageR::NormalizeObject(res_slic$slic_data)
plot_slic = grDevices::as.raster(plot_slic)
graphics::plot(plot_slic)

plot_slico = OpenImageR::NormalizeObject(res_slico$slic_data)
plot_slico = grDevices::as.raster(plot_slico)
graphics::plot(plot_slico)
```

Try more advanced methods from `SuperpixelImageSegmentation`. (Need to run the image commands in the console to work)


```{r}
init = Image_Segmentation$new()

spx = init$spixel_segmentation(input_image = im, 
                               superpixel = 600, 
                               AP_data = TRUE,
                               use_median = TRUE, 
                               sim_wL = 3, 
                               sim_wA = 10, 
                               sim_wB = 10,
                               sim_color_radius = 10, 
                               verbose = TRUE)

imageShow(spx$AP_image_data)
```

```{r}
spx_km = init$spixel_segmentation(input_image = im, 
                                  superpixel = 600, 
                                  AP_data = TRUE,
                                  use_median = TRUE, 
                                  sim_wL = 3, 
                                  sim_wA = 10, 
                                  sim_wB = 10,
                                  sim_color_radius = 10, 
                                  kmeans_method = "kmeans",
                                  kmeans_initializer = "kmeans++",
                                  kmeans_num_init = 3, 
                                  kmeans_max_iters = 100,
                                  verbose = TRUE)

imageShow(spx_km$AP_image_data)
```

```{r}
spx_mbkm = init$spixel_segmentation(input_image = im, 
                                    superpixel = 600, 
                                    AP_data = TRUE,
                                    use_median = TRUE, 
                                    sim_wL = 3, 
                                    sim_wA = 10, 
                                    sim_wB = 10,
                                    sim_color_radius = 10, 
                                    kmeans_method = "mini_batch_kmeans",
                                    kmeans_initializer = "kmeans++",
                                    kmeans_num_init = 3, 
                                    kmeans_max_iters = 100,
                                    minib_kmeans_batch = 10, 
                                    minib_kmeans_init_fraction = 0.75,
                                    verbose = TRUE)

imageShow(spx_mbkm$KMeans_image_data)
```


